---
title: "State TaxData Project"
author: Don Boyd
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook: 
    df_print: paged
    toc: yes
    toc_depth: 5
    number_sections: true
editor_options: 
  chunk_output_type: console
---

# Project Statement of Work
See [this](https://docs.google.com/document/d/1K-3V3Acfeoyc5VvcPNnfsFkUq37fgMQEOSmR3_yGENA/edit?usp=sharing). Also see the separate rmd and md file on the OSPC agreement.

The primary goals of this project are to (1) adapt and extend methods and tools developed in previous incubator projects, allowing us to develop state-specific income tax microdata files from the existing IRS SOI national Public Use File (PUF), and (2) develop at least one preliminary file for one state.


# Yimeng's Project Statement of Work
> Augment Boyd’s current project in which he is developing methods for creating a single state microdata tax file, in two ways: (a) rigorously compare file quality under several different approaches under investigation, and (b) port the approach that produces the best results from R to Python so that it can readily be integrated with TaxData.


<!-- First set things up, then run things -->


# Set things up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r system_specific_info, include=FALSE}
# change these as needed
# pufdir <- "C:/Users/donbo/Dropbox/OSPC - Shared/IRS_pubuse_2011/"

```


```{r loads}
source(here::here("r/includes", "libraries.r"))
source(here::here("r/includes", "globals.r"))
globals
source(here::here("r/includes", "functions_state_puf.r"))

ns <- function(df) {names(df) %>% sort}

```



# Outline of work

1. ONETIME:
    + Get and save 2011 targets for all states
    + Get and save an rds version of the PUF, with aggregate records excluded
    + Get and save ratio-adjusted weights for all states



# Steps

## Get names of essential PUF variables, based on synthetic files and what we learned after
```{r ONETIME_vnames, eval=FALSE}
globals
df <- read_csv("C:/Users/donbo/Google Drive/synpuf/syntheses/synpuf20_no_disclosures.csv", n_max=0)
ns(df)
# variables needed to run the file through tax calculator -- see note below
# E03260  Deduction for self-employment tax
# XOCAH   Exemptions for Children Living at Home
# XOCAWH  Exemptions for Children Living Away from Home
# XOODEP  Exemptions for Other Dependents 
# XOPAR   Exemptions for Parents Living at Home or Away from Home

# we will get other variables after we run the "grown" file through Tax-Calculator
#   c00100 --> E00100
#   ?? --> E02500
#   ?? --> E04800
needed_puf_vars <- c(names(df), "E03260", "XOCAH", "XOCAWH", "XOODEP", "XOPAR")
saveRDS(needed_puf_vars, here::here("data", "needed_puf_vars.csv"))

# readRDS(here::here("data", "needed_puf_vars.csv"))

# On Mon, Nov 4, 2019 at 10:03 AM Yin, Yimeng <yyin@albany.edu> wrote:
# Hi Don, 
# 
# I found I still can't run the synthetic puf through taxdata after filling the missing variables. 
# 
# There are 8 missing variables in the synthetic puf:
# 
# puf['xocah']  "Exemptions for Children Living at Home"
# puf['xocawh'] "Exemptions for Children Living Away from Home"
# puf['xoodep']  "Exemptions for Other Dependents"
# puf['xopar']    "Exemptions for Parents Living at Home or Away from Home"
# puf['e03260'] "Deduction for self-employment tax"
# puf['e00100']  "Adjusted Gross Income (deficit)  (AGI)  (+/-)"
# puf['e02500'] "Social Security benefits in AGI"
# puf['e04800'] "Taxable income"

```



## Get and save targets for all states

```{r ONETIME_targets, eval=FALSE}
# create a clean file of 2011 SOI Historical Table 2 targets, all states, saved in data/hist2_2011.csv
source(here::here("r", "0a_prepare_soi2011_targets_csvformat.r"))
save_soi2011_targets_csvformat(globals) # creates ./data/hist2_2011.rds with soi targets all states
# readRDS(here::here("data", "hist2_targets2011.rds"))

source(here::here("r", "0b_prepare_soicsv_targets.r"))
prepare_soicsv_targets_long(globals, 2017) # creates ./data/hist2_2011.rds with soi targets all states
# readRDS(here::here("data", paste0("hist2_targets", 2017, ".rds")))

# puf <- get_puf_xagg() # ORIGINAL PUF less the 4 aggregate records, with a wt variable
# glimpse(puf)

# prepare 

# prepare puf for tax-calculator and get 2017 agi

# Prepare a file for each year with RECID and ratio-adjusted weights for EACH state
source(here::here("r", "0b_prepare_ratio_adjusted_state_weights.r"))


```

