---
title: "State TaxData Project"
author: Don Boyd
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook: 
    df_print: paged
    toc: yes
    toc_depth: 5
    number_sections: true
editor_options: 
  chunk_output_type: console
---

# Project Statement of Work
See [this](https://docs.google.com/document/d/1K-3V3Acfeoyc5VvcPNnfsFkUq37fgMQEOSmR3_yGENA/edit?usp=sharing). Also see the separate rmd and md file on the OSPC agreement.

The primary goals of this project are to (1) adapt and extend methods and tools developed in previous incubator projects, allowing us to develop state-specific income tax microdata files from the existing IRS SOI national Public Use File (PUF), and (2) develop at least one preliminary file for one state.


# Yimeng's Project Statement of Work
> Augment Boyd’s current project in which he is developing methods for creating a single state microdata tax file, in two ways: (a) rigorously compare file quality under several different approaches under investigation, and (b) port the approach that produces the best results from R to Python so that it can readily be integrated with TaxData.


<!-- First set things up, then run things -->


# Set things up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r system_specific_info, include=FALSE}
# change these as needed
# pufdir <- "C:/Users/donbo/Dropbox/OSPC - Shared/IRS_pubuse_2011/"

```


```{r loads}
source(here::here("r/includes", "libraries.r"))
source(here::here("r/includes", "globals.r"))
globals
source(here::here("r/includes", "functions_state_puf.r"))
source(here::here("r/includes", "functions_ipopt_djb.r"))

```



# Outline of work
[TO COME]


# Steps

## Get names of essential PUF variables, based on synthetic files and what we learned after
```{r ONETIME_vnames, eval=FALSE}
globals
df <- read_csv("C:/Users/donbo/Google Drive/synpuf/syntheses/synpuf20_no_disclosures.csv", n_max=0)
ns(df)
# variables needed to run the file through tax calculator -- see note below
# E03260  Deduction for self-employment tax
# XOCAH   Exemptions for Children Living at Home
# XOCAWH  Exemptions for Children Living Away from Home
# XOODEP  Exemptions for Other Dependents 
# XOPAR   Exemptions for Parents Living at Home or Away from Home

# we will get other variables after we run the "grown" file through Tax-Calculator
#   c00100 --> E00100  AGI
#   c02500 --> E02500  Social Security benefits in AGI
#   c04800 --> E04800  Regular taxable income
needed_puf_vars <- c(names(df), "E03260", "XOCAH", "XOCAWH", "XOODEP", "XOPAR")
saveRDS(needed_puf_vars, here::here("data", "needed_puf_vars.csv"))
# readRDS(here::here("data", "needed_puf_vars.csv"))

# On Mon, Nov 4, 2019 at 10:03 AM Yin, Yimeng <yyin@albany.edu> wrote:
# Hi Don, 
# 
# I found I still can't run the synthetic puf through taxdata after filling the missing variables. 
# 
# There are 8 missing variables in the synthetic puf:
# 
# puf['xocah']  "Exemptions for Children Living at Home"
# puf['xocawh'] "Exemptions for Children Living Away from Home"
# puf['xoodep']  "Exemptions for Other Dependents"
# puf['xopar']    "Exemptions for Parents Living at Home or Away from Home"
# puf['e03260'] "Deduction for self-employment tax"
# puf['e00100']  "Adjusted Gross Income (deficit)  (AGI)  (+/-)"
# puf['e02500'] "Social Security benefits in AGI"
# puf['e04800'] "Taxable income"

# See user guide: http://pslmodels.github.io/Tax-Calculator/uguide.html

```


# prepare state targets for 2011 and 2017
```{r ONETIME_targets, eval=FALSE}
# NOTE: I put all dollar targets into actual dollars, rather than $ thousands as they are in source data
# Later we may do some scaling in optimization routines, howev er

# create a clean file of 2011 SOI Historical Table 2 targets, all states, saved in data/hist2_2011.csv
source(here::here("r", "0a_prepare_soi2011_targets_csvformat.r"))
save_soi2011_targets_csvformat(globals) # creates ./data/hist2_2011.rds with soi targets all states
# readRDS(here::here("data", "hist2_targets2011.rds"))

source(here::here("r", "0b_prepare_soicsv_targets.r"))
prepare_soicsv_targets_long(globals, 2017) # creates ./data/hist2_2011.rds with soi targets all states
# readRDS(here::here("data", paste0("hist2_targets", 2017, ".rds")))

```


# prepare a national PUF file with weights for 2017
```{r ONETIME_puf2017_national}
# get 2011 to 2017 growth ratios for SOI - we use these to help calculate grow factors - just for US
source(here::here("r", "0c_get_soi_historical2_growthratios.r"))
# readRDS(here::here("data", "growthratios_2011to2017.rds"))

# prepare growth factors based on the SOI growth rates and other considerations
source(here::here("r", "0c_prepare_growfactors.r"))
# readRDS(here::here("data", "grow_factors_final_2011to2017.rds")) # grow_factors_final

# "grow" the file to 2017 and prepare puf for tax-calculator and get 2017 agi
source(here::here("r", "0d_grow_puf.r"))
# readRDS(paste0(globals$statedir, "puf2017_unweighted.rds"))

# two steps:
#   1) run file through tax-calculator to add agi2017 and certain other calculated variables
#   2) construct an initial 2017 US weight that hits the US agi target
source(here::here("r", "0e_get_agi2017.r"))
# readRDS(paste0(globals$statedir, "puf2017_weighted.rds")) # this is the national file for 2017

```


# construct ratio-adjusted weights for all states, and save them -- a starting point for optimization
```{r ONETIME_state_ratio_adjusted_weights}
# mechanically construct ratio-adjusted state weights for the state in question, as a starting point for optimization
source(here::here("r", "0f_get_ratioweights2017.r"))
# readRDS(paste0(globals$statedir, "state_ratio_adjusted_weights_2017.rds")) # long file, a weight for every state

```


# Optimize for a single state
```{r optimize_single_state}
# reweight the file by 2017 AGI bins to hit or come close to state targets (Ipopt, python can be CVXOPT)
# NOT READY: source(here::here("r", "0g_reweight_toward_targets.r"))  
# readRDS(here::here("results", "NY_x.rds"))

# evaluate results



```


# Analyze results
```{r}
xvalues <- readRDS(here::here("results", "NY_x.rds"))
# pufbase_state <- readRDS(paste0(globals$statedir, "pufbase_state.rds"))
pufbase_all <- readRDS(paste0(globals$statedir, "puf2017_weighted.rds"))
ns(pufbase_all)

comp <- pufbase_all %>%
  left_join(xvalues, by="RECID") %>%
  mutate(weight_state=wt_init * x) %>%
  rename(E00100=c00100, E09600=c09600, E05800=c05800) 
glimpse(comp)
ns(comp)

comp %>% select(-wt_init, -x, c62100) %>% write_csv(paste0(globals$state_puf_shared, "puf2017_usny.csv"))


# nystats <- readRDS(here::here("data", paste0("hist2_targets", 2017, ".rds"))) %>%
#   filter(stabbr=="NY")

nystats <- readRDS(here::here("data", "targets_all.rds")) %>%
  filter(stabbr=="NY")

compvars <- c("E00100", "E00200", "E00300", "E02300", "E05800", "E09600", "E18500")

tab <- comp %>%
  select(RECID, AGI_STUB, weight_state, compvars) %>%
  group_by(AGI_STUB) %>%
  summarise_at(vars(compvars), list(A=~sum(. * weight_state), N=~sum(weight_state * (. != 0)))) %>%
  pivot_longer(-AGI_STUB) %>%
  mutate(name=case_when(str_sub(name, -1) == "A" ~ paste0("A", str_sub(name, 2, 6)),
                        str_sub(name, -1) == "N" ~ paste0("N", str_sub(name, 2, 6)))) %>%
  left_join(nystats %>% select(name=h2vname, AGI_STUB, target, table_desc))

# df <- tab
# vname <- "N00100"

make_flex <- function(vname, df) {
    # define scaling and header text
  htext <- df %>% filter(name==vname) %>% .$table_desc %>% unique
  if(str_sub(vname, 1, 1)=="A") {
    type <- "amount"
    scale <- 1e9
    htext <- paste0(htext, "\n(Billions of dollars)")
  } else if(str_sub(vname, 1, 1)=="N") {
    type=="number"
    scale <- 1000
    htext <- paste0(htext, "\n(Thousands of returns)")
  }
  
  tabdf <- df %>%
    filter(name==vname) %>%
    arrange(AGI_STUB) %>% 
    adorn_totals() %>%
    mutate(AGI_STUB=ifelse(AGI_STUB=="Total", 0, AGI_STUB),
           AGI_STUB=as.integer(AGI_STUB),
           AGI_STUB=factor(AGI_STUB, levels=globals$agilabs$AGI_STUB, labels=globals$agilabs$agi_label)) %>%
    mutate(calc=value / scale,
           target=target / scale,
           diff=calc - target,
           pdiff=diff / target * 100) %>%
    select(AGI_STUB, name, table_desc, target, calc, diff, pdiff)
  
  ft <- tabdf %>%
    flextable(col_keys = c("AGI_STUB", 
                         "separator1",
                         "target", "calc", "diff",
                         "separator2",
                         "pdiff")) %>%
  set_header_labels(AGI_STUB="AGI range",
                    target = "Historical Table 2",
                    calc = "Calculated\nvalue",
                    diff="Difference",
                    pdiff="% Difference") %>%
  add_header_row(values=c("", "", htext, "", ""), colwidths=c(1, 1, 3, 1, 1)) %>%
  align(i = 1, j = NULL, align = "center", part = "header") %>%
  colformat_num(j = c("target", "calc", "diff"),
                big.mark=",", digits = 1, na_str = "N/A") %>%
  colformat_num(j = c("pdiff"),
                big.mark=",", digits = 1, na_str = "N/A") %>%
  width(j= ~ AGI_STUB, width=3) %>%
  width(j = ~ target + calc + diff + pdiff, width=1.5) %>%
  width(j = ~ separator1 + separator2, width = .3) %>%
  height_all(part = "header", height = .4) %>%
  height_all(part = "body", height = .4) %>%
  # now control the themes # theme_booktabs() %>%
  border_remove() %>%
  hline(i=1, j=~target, border = fp_border(width = 1), part = "header") %>%
  hline_bottom(j= ~ AGI_STUB + target + calc + diff + pdiff, border = fp_border(width = 1), part = "header") %>%
  hline(i=10, j= ~ AGI_STUB + target + calc + diff + pdiff, border = fp_border(width = 1), part = "body") %>%
  fontsize(size = 20, part = "all") %>%
  fontsize(i=1, size = 20, part = "header")
  
  return(ft)  
}


# A00100 <- make_flex("A00100", tab)
# save_as_image(A00100, path = here::here("results", "A00100.png"), zoom=7)

savetab <- function(vname, tab) {
  ftab <- make_flex(vname, tab)
  save_as_image(ftab, path = here::here("results", paste0(vname, ".png")), zoom=7)
}

A02300

savetab("A00100", tab)
savetab("A00200", tab)
savetab("A00300", tab)
savetab("A02300", tab)
savetab("A05800", tab)
savetab("A09600", tab)
savetab("A18500", tab)

savetab("N00100", tab)
savetab("N00200", tab)
savetab("N00300", tab)
savetab("N02300", tab)
savetab("N05800", tab)
savetab("N09600", tab)
savetab("N18500", tab)

```



