---
title: "State TaxData Project"
author: Don Boyd
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_notebook: 
    df_print: paged
    toc: yes
    toc_depth: 5
    number_sections: true
editor_options: 
  chunk_output_type: console
---

# Project Statement of Work


# Yimeng's Project Statement of Work
> Augment Boyd’s current project in which he is developing methods for creating a single state microdata tax file, in two ways: (a) rigorously compare file quality under several different approaches under investigation, and (b) port the approach that produces the best results from R to Python so that it can readily be integrated with TaxData.


<!-- First set things up, then run things -->


# Set things up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r system_specific_info, include=TRUE}
# change these as needed
# pufdir <- "C:/Users/donbo/Dropbox/OSPC - Shared/IRS_pubuse_2011/"

```


```{r loads}
source(here::here("r/includes", "libraries.r"))
source(here::here("r/includes", "globals.r"))
globals
source(here::here("r/includes", "functions_state_puf.r"))

```



# Outline of work

1. Create 



# Steps

## Create file

```{r ONETIME}
# create a clean file of 2011 SOI Historical Table 2 targets, all states, saved in data/hist2_2011.csv
source(here::here("r", "0a_get_and_save_SOI_state_summary_files.r"))

source(here::here("r", "0b_prepare_ratio_adjusted_state_weights.r"))




```

